---
title: "Introduction to regmorbidity"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to regmorbidity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(regmorbidity)
```

## What this package does

regmorbidity derives **condition indicators with dates** from Danish register
data — hospital diagnoses (LPR) and dispensings (LMDB) — using editable CSV
code lists rather than hard-coded rules. It computes no score and is not an
implementation of any published multimorbidity index.

The public surface is **15 exported functions**. Prefer the happy path below;
backup / QA / advanced / provisional functions are listed at the end of this
section and used only where noted.

**Happy path (loud)**

1. `mb_codelist` — load/filter lists (`conditions=`, `vocab=`)
2. `mb_extract_diagnosis` — LPR onset
3. `mb_extract_medication_batch` — preferred medication onset (SQL, low RAM)
4. `mb_merge_all` — combine dx+rx extract directories → long
5. `mb_to_wide` → `mb_count_conditions` — ever-after counts as of a date

**Backup / special:** `mb_extract_medication` (sequential; `keep_events` /
one-condition debug), `mb_merge_conditions` (one condition in memory),
`mb_load_conditions` (one extract outdir → long; medication-only),
`mb_prevalence` (lookback on **raw events**, not onset `.rds`).

**QA:** `mb_check_codes`, `mb_lookup`, `mb_overlap`.

**Advanced:** `mb_inspect_codes` — code column string lengths in the register
(prefilter debugging only; not list review).

**Provisional:** `mb_apply_exclusions` — optional stage-2; incomplete vs Prior.

Output is onset-only by default. Design notes
(`ASSUMPTIONS_AND_LIMITATIONS.txt`, `DECISIONS.md`) live in the source working
tree if present; otherwise use this vignette and `?regmorbidity`.

## What you supply

1. **LPR** as **one** combined table: person id, ICD code, contact date
   (defaults: `pnr`, `C_DIAG`, `D_INDDTO`). The package does not merge LPR2,
   LPR3, or psychiatric tables for you.
2. **LMDB**: person id, **full** ATC, dispensing date (defaults: `pnr`,
   `atc`, `eksd`). An optional short ATC column (e.g. `atc2`) is used only by
   the sequential medication extractor as a pushed-down prefilter.
3. **A code list** via `mb_codelist()` — bundled lists are **ATC + first-pass
   ICD** (distress archived out of the default set). Filter with
   `conditions=` and/or `vocab=` (e.g. `vocab = "ICD10"` for diagnosis,
   `vocab = "ATC"` for medication).

Date windows use inclusive `from` / `to`. On LMDB medication extractors,
`from` is **required** (error if omitted/`NULL`); `to` is optional. Prefer
`from = as.Date("1997-01-01")` unless you have handled the pre-1997
mother-CPR artifact another way. Diagnosis extract may still omit `from`
(entire register).

### Preflight (optional, before a long DST run)

Lead with whether list codes hit anything in the register:

```{r, eval = FALSE}
mb_check_codes(lpr, dx_codes)
mb_check_codes(lmdb, rx_codes)
mb_lookup("C02CA01")          # which conditions claim this code?
mb_overlap(vocab = "ATC")     # codes shared between conditions
```

#### Advanced: column lengths

`mb_inspect_codes()` reports **code column string lengths** in the register
(e.g. is `atc2` 3 chars?). Only needed before sequential extract / prefilter
debugging. Not list review.

```{r, eval = FALSE}
mb_inspect_codes(lmdb)
mb_inspect_codes(lpr)
```

## Happy path

Extractors take the code list and write one onset `.rds` per condition.
Diagnoses first, then medications (batch), then merge.

```{r}
# Tiny synthetic registers for this vignette only
lpr <- data.frame(
  pnr      = c("1", "2", "3"),
  C_DIAG   = c("DI10", "DI10", "DE11"),
  D_INDDTO = as.Date(c("2008-01-01", "2009-06-01", "2011-01-01")),
  stringsAsFactors = FALSE
)
lmdb <- data.frame(
  pnr  = c("1", "1", "2", "2", "3"),
  atc  = c("C09AA02", "C09AA02", "C09AA02", "C09AA02", "A10BA02"),
  eksd = as.Date(c("2010-01-01", "2010-03-01",
                   "2011-01-01", "2011-04-01", "2012-01-01")),
  stringsAsFactors = FALSE
)
```

```{r, eval = FALSE}
dx_codes <- mb_codelist(conditions = "hypertension", vocab = "ICD10")
rx_codes <- mb_codelist(vocab = "ATC")

mb_extract_diagnosis(
  lpr,
  codes  = dx_codes,
  outdir = "data/dx",
  from   = as.Date("1995-01-01")
)

mb_extract_medication_batch(
  lmdb,
  codes  = rx_codes,
  outdir = "data/rx",
  from   = as.Date("1997-01-01")
)

# Merge returns a long data frame. mb_load_conditions() is for one extract
# outdir when you are not merging (e.g. medication-only).
long <- mb_merge_all("data/dx", "data/rx")
wide <- mb_to_wide(long)
mb_count_conditions(wide, as_of = "2015-01-01")
```

### Worked miniature (in memory, one condition)

Uses `mb_merge_conditions()` — the in-memory, one-condition backup — so the
vignette stays file-free and evaluable:

```{r}
dx_codes <- mb_codelist(data.frame(
  condition = "hypertension", vocab_id = "ICD10", code = "I10",
  exclude = FALSE, stringsAsFactors = FALSE
))
rx_codes <- mb_codelist(data.frame(
  condition = "hypertension", vocab_id = "ATC", code = "C09",
  exclude = FALSE, min_prescriptions = 2L, window_days = 365L,
  stringsAsFactors = FALSE
))

dx <- mb_extract_diagnosis(lpr, codes = dx_codes, outdir = NULL,
                           verbose = FALSE)
rx <- mb_extract_medication_batch(lmdb, codes = rx_codes, outdir = NULL,
                                  from = as.Date("1997-01-01"),
                                  verbose = FALSE)

merged <- mb_merge_conditions(dx$hypertension, rx$hypertension)
wide <- mb_to_wide(merged)
mb_count_conditions(wide, as_of = "2015-01-01")
```

For a study cohort rather than everyone in the register, pass `ids` on the
extract call. That is a join against the lazy connection, not a written-out
id list.

```{r, eval = FALSE}
mb_extract_medication_batch(lmdb, codes = rx_codes, outdir = "data/rx",
                            from = as.Date("1997-01-01"),
                            ids = cohort$pnr)
```

## Diagnosis and merge

`mb_extract_diagnosis()` needs no prescription rule: one matching diagnosis
is the condition, and onset is the earliest date. Feed it **one** already-
combined LPR table. Prefer ICD rows via
`mb_codelist(conditions = ..., vocab = "ICD10")`.

`"OR"` (default) in merge takes whichever of diagnosis / medication came
first; `"AND"` requires both and takes the later date. Which applies can
come from an optional `logic` column in the code list.
`mb_merge_all()` walks a pair of extract directories (happy path).
`mb_merge_conditions()` merges one condition’s two data frames in memory.

## Medication rules in brief

Medication conditions usually need more than one dispensing. The code list
carries `min_prescriptions` and `window_days` (often 2 within 365 days).
Same-day dispensings are de-duplicated by default
(`dedupe_same_day = TRUE`), so two packages on one pharmacy visit count as
one event. Leaving `window_days` as `NA` means any distance apart.

Three settings that change counts:

- **`from` / `to`** — inclusive window. On LMDB medication extractors
  `from` is required; `to` optional (`NULL` = no upper bound). Prefer
  `from = as.Date("1997-01-01")` on LMDB (mother-CPR artifact).
- **`dedupe_same_day`** — default `TRUE`.
- **`window_days` / `min_prescriptions`** — from the code list, overridable
  on the extract call where the API allows.

## Ever/never, or a lookback (batch ≠ events)

`mb_count_conditions()` counts a condition from its onset date on: once met,
always met (as of the date you pass). State that estimand wherever you
report a number.

If you need "still evidence within the last N days", use `mb_prevalence()`
on **raw matching records**, not onset-only `.rds`.

**Batch does not collect raw rows.** `mb_extract_medication_batch()` writes
onset only. To keep events for prevalence (or one-condition debug), use the
sequential extractor:

```{r}
outdir <- tempfile()
mb_extract_medication(lmdb, codes = rx_codes, outdir = outdir,
                      from = as.Date("1997-01-01"),
                      keep_events = TRUE, verbose = FALSE)
disp <- readRDS(file.path(outdir, "hypertension_all_events.rds"))

mb_prevalence(disp, as_of = "2011-01-01", lookback = 365,
             min_prescriptions = 2, window_days = 365)
```

- `keep_events = TRUE` writes `<condition>_all_events.rds` beside the onset
  file. Batch has no equivalent.
- With `outdir = NULL`, side files are not written.
- `lookback = Inf` matches `mb_count_conditions(as_of = )`.
- `mb_load_conditions()` skips `*_all_events.rds` automatically.

Prefer the full-list / batch path. Switch to sequential when you need
`keep_events`, or when the population / year range makes a full-list pass
awkward. Both write the same onset `.rds` shape, so merge / load / wide /
count stay unchanged.

## Code-list overlaps (QA)

Some bundled ATC ranges nest (e.g. hypertension `C02C` and prostate
`C02CA`). An overlap is not automatically wrong, but one dispensing
creating two conditions should be a deliberate choice:

```{r, eval = FALSE}
mb_overlap()              # all vocabs in the list
mb_overlap(vocab = "ATC")
mb_lookup("C02CA01")      # which conditions claim a code
```

## Optional: Prior exclusion rules (provisional)

Extraction does **not** apply Prior's cross-condition exclusions. After you
have a long or wide onset table, you can optionally call
`mb_apply_exclusions()` — provisional; incomplete vs Prior; not on the happy
path. Distress was dropped from the default lists, so do not lean on distress
arms here. Hypertension still has gaps around C03 / heart failure / CKD
relative to a full Prior-style exclusion set.

```{r, eval = FALSE}
# long from mb_merge_all() / mb_load_conditions(), or wide from mb_to_wide()
long2 <- mb_apply_exclusions(long, as_of = as.Date("2015-01-01"))
# timing: "as_of" (default when as_of given), "onset_order", or "ever_ever"
```

## Where to read more

`?regmorbidity` — all 15 exports in the same tiers.

In the source working tree if present (often gitignored / local):

- `ASSUMPTIONS_AND_LIMITATIONS.txt` — short checklist before reporting a number
- `DECISIONS.md` — design choices, sources, and costs
- `METHODS.md` — article-ready prose
