---
title: "Introduction to regmorbidity"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to regmorbidity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(regmorbidity)
```

## What this vignette is for

This vignette **deepens** the README. Read the README first for install, what
you supply, the DST DuckDB path, the recommended five-step workflow, and the
core function list. This page does **not** repeat a full API inventory.

Register chunks below are not executed. On DST, `lpr` and `lmdb` are your
register connections (see the README). Load lists with `mb_codelist()`.

regmorbidity derives **condition indicators with dates** from Danish register
data - hospital diagnoses (LPR) and dispensings (LMDB) - using editable CSV
code lists rather than hard-coded rules. It computes no score and is not an
implementation of any published multimorbidity index. The bundled lists take
their starting point in Prior et al. (2016) and are a starting point to
revise, not an instrument to cite.

## Directory form (real DST-style outdirs)

Same pipeline as the README, writing one `.rds` per condition. Use a
**fresh** `outdir`, or clear old files first: with `resume = TRUE` (the
default), extractors skip conditions whose `.rds` already exists. Reusing a
dirty folder of leftover results can silently keep stale onsets.

```{r, eval = FALSE}
# On DST: lpr and lmdb are your registers (see README).
# All bundled conditions:  codes <- mb_codelist()
# A named subset:          codes <- mb_codelist(conditions = c("hypertension", "diabetes"))
# Your own folder of CSVs: codes <- mb_codelist("path/to/my_codelists/")
codes <- mb_codelist(conditions = "hypertension")

mb_extract_diagnosis(
  lpr, codes = codes, outdir = "data/dx",
  from = as.Date("1995-01-01"),  # inclusive start; placeholder
  to   = as.Date("2018-12-31")   # optional; omit to leave open-ended
)
mb_extract_medication_batch(
  lmdb, codes = codes, outdir = "data/rx",
  from = as.Date("1997-01-01"),  # required on LMDB (mother-CPR; see README)
  to   = as.Date("2018-12-31")   # optional; not required
)

long <- mb_merge_all("data/dx", "data/rx")
wide <- mb_to_wide(long)
mb_count_conditions(wide, as_of = "2015-01-01")  # placeholder as-of date
```

Medication-only studies (no diagnosis half) can load one outdir directly:

```{r, eval = FALSE}
# Skips *_all_events.rds automatically if any are present
long_rx_only <- mb_load_conditions("data/rx")
```

## Restricting to a cohort (`ids`)

For a study cohort rather than everyone in the register, pass `ids` on the
extract call. That is a join against the lazy connection, not a written-out
id list dumped to disk.

```{r, eval = FALSE}
# On DST: lmdb is your LMDB connection. cohort$pnr is your study ids.
codes <- mb_codelist(conditions = "hypertension")

mb_extract_medication_batch(
  lmdb, codes = codes, outdir = "data/rx",
  from = as.Date("1997-01-01"),  # required on LMDB
  to   = as.Date("2018-12-31"),  # optional; not required
  ids  = cohort$pnr
)
```

## Batch vs sequential detail

Prefer **batch** (`mb_extract_medication_batch`) for medication onset: one SQL
pass, low RAM, same onset `.rds` shape as sequential.

Switch to **sequential** (`mb_extract_medication`) when:

1. you need **raw events** for prevalence (`keep_events = TRUE`), or
2. you want one-condition disk checkpoints / resume on an awkward cohort or
   year range.

**Why both exist.** Batch never materialises matching dispensings in R; it asks
DuckDB for the onset only - that is the right default for onset studies.
Sequential walks conditions (with an optional short `atc2` prefilter), and with
`keep_events = TRUE` it also writes `<condition>_all_events.rds` beside the
onset file. Batch has no equivalent side file, so prevalence cannot be fed from
a batch-only outdir. Merge / load / wide / count stay the same either way
because both write the same onset `.rds` shape.

## Medication rules in brief

Medication conditions usually need more than one dispensing. The code list
carries `min_prescriptions` and `window_days` (often 2 within 365 days).
Same-day dispensings are de-duplicated by default
(`dedupe_same_day = TRUE`). Leaving `window_days` as `NA` means any distance
apart.

Three settings that change counts:

- **`from` / `to`** - inclusive. On LMDB, `from` is required; prefer
  `from = as.Date("1997-01-01")` because of the mother-CPR recording change
  (Pottegard et al. 2016; see README). `to` is optional; omit it to leave the
  upper end open.
- **`dedupe_same_day`** - default `TRUE`.
- **`window_days` / `min_prescriptions`** - from the code list.

## Prevalence on raw events

`mb_count_conditions()` is ever-after from onset. Use `mb_prevalence` when you
need "still evidence within the last N days" on **raw matching records**, not
onset-only `.rds`. That requires sequential extract with `keep_events = TRUE`.

```{r, eval = FALSE}
# On DST: lmdb is your LMDB connection.
# Sequential extract writes <condition>_all_events.rds beside the onset file.
codes <- mb_codelist(conditions = "hypertension")

mb_extract_medication(
  lmdb, codes = codes, outdir = "data/rx",
  from = as.Date("1997-01-01"),  # required on LMDB
  to   = as.Date("2018-12-31"),  # optional; not required
  keep_events = TRUE
)
disp <- readRDS(file.path("data/rx", "hypertension_all_events.rds"))

mb_prevalence(
  disp, as_of = "2011-01-01", lookback = 365,
  min_prescriptions = 2, window_days = 365
)
```

- `keep_events = TRUE` writes `<condition>_all_events.rds`. Batch has no
  equivalent.
- With `outdir = NULL`, side files are not written.
- `lookback = Inf` matches `mb_count_conditions(as_of = )`.

## Merge logic (OR / AND)

`"OR"` (default) takes whichever of diagnosis and medication came first.
`"AND"` requires both and takes the later date. An optional `logic` column
in the code list sets this per condition. `mb_merge_all()` applies that rule
across a pair of extract directories. `mb_merge_conditions()` does the same
for one condition's two in-memory data frames (the diagnosis half and the
medication half).

## Optional: Prior exclusion rules (provisional)

Extraction does **not** apply Prior's cross-condition exclusions.

Use `mb_apply_exclusions` when you already have a long or wide onset table and
want an optional stage-2 pass. This is **provisional / first-pass**: incomplete
versus Prior, and not part of the recommended five-step workflow. Distress was
dropped from the default lists. Hypertension still has gaps around C03 / heart
failure / CKD.

Timing options (pass via `timing=`):

- **`"as_of"`** - default when `as_of` is given: apply exclusions relative to
  that date.
- **`"onset_order"`** - order by onset dates among conditions present.
- **`"ever_ever"`** - if both conditions ever appear, apply the exclusion
  regardless of timing.

```{r, eval = FALSE}
# long from mb_merge_all() / mb_load_conditions(), or wide from mb_to_wide()
long2 <- mb_apply_exclusions(long, as_of = as.Date("2015-01-01"))
# timing: "as_of" (default when as_of given), "onset_order", or "ever_ever"
```

## Advanced: code-list QA and inspect

These helpers are for checking lists and register columns before a long DST
run. They are not part of the recommended five-step workflow.

**`mb_check_codes`** - for each code in your list, count how many register rows
match (prefix match, same as the extractors). Typos, retired codes, or a
missing ATC chapter otherwise give **silent zeros**.

```{r, eval = FALSE}
# On DST: lmdb is your LMDB connection.
# One row per list code, with how many register rows match (prefix).
codes <- mb_codelist(conditions = "hypertension")
mb_check_codes(lmdb, codes, code_col = "atc")
```

A zero here means the extractor would also return empty for that code - fix
the list (or the register window) before a multi-hour DST run.

```{r, eval = FALSE}
mb_lookup("C02CA01")          # which conditions claim this code?
mb_overlap(vocab = "ATC")     # codes shared between conditions
```

**`mb_inspect_code_lengths`** - reports string lengths of code columns in the
*register* sample (e.g. whether `atc2` is 3 characters). Use before sequential
medication extract / prefilter debugging. This is **not** for reviewing the
code list; that is `mb_check_codes`.

Why it exists: the sequential path can push a coarse filter on `atc2` with
`prefilter_len = 3`. If `atc2` is not 3 characters (or is missing), that
prefilter silently matches nothing.

```{r, eval = FALSE}
# On DST: lmdb is your LMDB connection.
mb_inspect_code_lengths(lmdb, cols = c("atc", "atc2"))
```

### Short failure modes

- **Wrong colnames** - defaults are LPR `pnr` / `C_DIAG` / `D_INDDTO` and LMDB
  `pnr` / `atc` / `eksd`. Pass `id_col`, `code_col`, and `date_col` when your
  tables differ.
- **Silent zeros** - codes that never match the register look like "nobody has
  this condition". Run `mb_check_codes` first.
- **Empty extract** - possible and valid when no rows qualify in the window;
  also the symptom of a bad code, bad date window, or wrong column name.
- **Dirty `outdir`** - do not reuse a folder of old `.rds` blindly; resume
  skips existing files.

## Where to read more

- **README** - install, DST path, recommended workflow, core functions, batch
  vs sequential overview.
- **`?regmorbidity`** - help for every exported function.

`ASSUMPTIONS_AND_LIMITATIONS.txt`, `DECISIONS.md`, and `METHODS.md` are
**local working-tree notes** (often gitignored). They are not part of a normal
GitHub install. If you installed from GitHub, use the README, this vignette,
and `?help`.
