---
title: "Introduction to regmorbidity"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to regmorbidity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(regmorbidity)
```

## What this package does, and does not do

regmorbidity derives **condition indicators with dates** from Danish register
data - dispensings (LMDB) and hospital diagnoses (LPR) - using editable CSV
code lists rather than code. It computes no score.

It is not an implementation of the Danish Multimorbidity Index or of any other
published index. The bundled code lists take their starting point in Prior et
al. (2016), and are meant to be revised, not cited as-is.

Everything in this vignette runs on small synthetic data. Nothing in the
package has been run against real register data - see
`ASSUMPTIONS_AND_LIMITATIONS.txt` in the source repository before reporting
any number from a real run.

## How this compares to Prior et al. (2016)

The bundled code lists take their starting point in the multimorbidity
algorithm published as a web supplement to Prior et al. (2016) - see
Reference, below. regmorbidity is not an implementation of that algorithm.
The differences are deliberate, and each is argued in full in `DECISIONS.md`
in the source repository; this is the short version.

**The same:**

- The starting point for which ATC/ICD-10 codes define each condition.
- The medication rule's window: 2 dispensings within 365 days (4 for pain).
  Prior's Stata, his Web Table 1, and the DST codebook all agree on 365 days
  independently - three sources, not one assumption carried over.
- Same-day dispensings collapsed to one event before counting, matching
  Prior's Stata exactly (`bys pid eksd: keep if _n == 1`) - though this one
  is itself provisional, see "What is still open" below: Jie's own script
  does not do this, so it is one of two joint decisions still open with her.
- Anchored prefix matching: the same semantics as
  `heaven::findCondition(match = 'start')`.

**Different, on purpose:**

- **Computes no score.** regmorbidity outputs condition indicators with
  dates, not a summed or weighted index. Prior himself adjusts for his 39
  conditions individually rather than summing them, and his own published
  supplement is titled "Algorithm for Physical and Mental Health
  Conditions", not a multimorbidity index.
- **Ever/never by default, not time-varying spells.** Prior's Stata splits
  follow-up into spells with recovery windows (2 years for most diagnoses,
  5 for cancer, 1 year for medication); once a condition starts here, it
  counts from then on. `mb_prevalence()`, described below, is a cheaper
  point-in-time alternative - still not spells.
- **Prior's composite exclusion rules are not implemented.** Three rules -
  dyslipidaemia only if not IHD, hypertension only if not IHD or heart
  failure, distress only if no other psychiatric diagnosis - reference
  *other* conditions and need a second pass over the merged table. Without
  them, several conditions are overcounted relative to a published number -
  the largest single source of divergence, and the more consequential of
  the two decisions still open with Jie.
- **Connective tissue uses D86 (sarcoidosis), not M86.** The one place this
  package knowingly differs from a direct instruction from Anders Prior:
  his Stata has `DM86`, the published Web Table 1 gives D86, and the
  surrounding code block is entirely the ICD-10 "systemic connective tissue
  disorders" range - a block sarcoidosis fits and osteomyelitis (M86) does
  not. Treated as a likely typo pending Anders's confirmation.
- **The ICD-10 half does not exist yet.** All 15 bundled code lists are
  ATC-only; every condition Prior defines partly or wholly by diagnosis is
  currently missing or understated, and 22 of his 39 conditions have no ATC
  side at all and are not in this package yet.
- **The code lists are meant to be edited, not cited as-is.** Prior's list
  is a starting point for a study to revise - see "Code lists", below - not
  a fixed instrument.

## Code lists

A code list is a data frame - or a folder of one CSV per condition - with one
row per code:

```{r}
codes <- data.frame(
  condition         = "hypertension",
  condition_label    = "Hypertension",
  vocab_id           = "ATC",
  code               = c("C02", "C03", "C07", "C08", "C09"),
  exclude            = FALSE,
  min_prescriptions  = 2,
  window_days        = 365,
  stringsAsFactors   = FALSE
)
codes <- mb_codelist(codes)
```

`mb_codelist()` validates on load and refuses a list that would fail quietly:
a condition whose rows disagree about the rule, a condition made only of
exclusions, duplicated codes. `min_prescriptions` and `window_days` describe
the *condition*, so every row of a condition must agree.

To start from the 15 ATC lists bundled with the package and edit from there:

```{r, eval = FALSE}
mb_write_codelist(mb_codelist(), "my_codelists")   # one file per condition
codes <- mb_codelist("my_codelists")               # read them back after editing
```

Before trusting a code list against a real register, two checks are worth
running first:

```{r, eval = FALSE}
mb_inspect_codes(lmdb)          # how long are the code columns, really
mb_check_codes(lmdb, codes)     # does every code match something at all
```

The second one is how a true zero is told from a typo or a missing chapter of
the register.

## Extracting medication conditions

The package needs nothing but an object `dplyr` can `filter()`, `select()`
and `collect()` on - a lazy arrow/DuckDB table on real data, or a plain data
frame here:

```{r}
lmdb <- data.frame(
  pnr  = c("1", "1", "1", "2"),
  atc  = c("C09AA01", "C09AA01", "C07AB02", "N02BE01"),
  atc2 = c("C09", "C09", "C07", "N02"),
  eksd = as.Date(c("2010-01-01", "2010-02-01", "2015-06-01", "2010-01-01")),
  year = c(2010, 2010, 2015, 2010),
  stringsAsFactors = FALSE
)

mb_extract_medication(lmdb, codes = codes, outdir = NULL, verbose = FALSE)
```

Person `1` has two hypertension dispensings 31 days apart - inside the
365-day window - so they meet the rule; person `2` has no matching codes at
all. With `outdir = NULL` the result comes back as a named list, one element
per condition, instead of being written to disk; **always check
`attr(result, "summary")$status`**, since a batch that finishes can still
contain per-condition errors.

For a real run, pass `outdir` so each condition is written to its own `.rds`
as soon as it is done, which makes an interrupted run resumable:

```{r, eval = FALSE}
mb_extract_medication(lmdb, codes = codes, outdir = "data/conditions")
long <- mb_load_conditions("data/conditions")   # one row per person-condition
wide <- mb_to_wide(long)                        # one column per condition
```

If the run is for a defined study cohort rather than everyone in the
register, pass `ids` to restrict before anything else runs:

```{r, eval = FALSE}
mb_extract_medication(lmdb, codes = codes, outdir = "data/conditions",
                      ids = cohort$pnr)
```

This is a join against the lazy connection, not a filter written out as a
list of ids, so it stays efficient with a large cohort on a large register.
It is a second, independent way to reduce how much ever needs to come home,
alongside the code-level filtering the two-stage match already does.

## Ever/never, or a point in time

`mb_count_conditions()` counts a condition from its onset date on: once met,
always met.

```{r}
long <- mb_extract_medication(lmdb, codes = codes, outdir = NULL,
                              verbose = FALSE)$hypertension
wide <- mb_to_wide(long)
mb_count_conditions(wide, as_of = "2020-01-01")
```

That is a different, cheaper estimand than a published index's recovery
windows, and it must be stated wherever a number is reported (see
`DECISIONS.md` 4.1 in the source repository for the full argument).

If instead you need "is there still evidence of this condition within the
last N days", `mb_prevalence()` applies the same rule to a record set
restricted to a lookback window. It needs the raw matching records, not the
extraction result, because `onset_date` alone cannot answer a lookback
question - which is why `save_dispensings = TRUE` needs a real `outdir`; with
`outdir = NULL` the raw records are not kept at all:

```{r}
outdir <- tempfile()
mb_extract_medication(lmdb, codes = codes, outdir = outdir,
                      save_dispensings = TRUE, verbose = FALSE)
disp <- readRDS(file.path(outdir, "hypertension_dispensings.rds"))

mb_prevalence(disp, as_of = "2011-01-01", lookback = 365,
             min_prescriptions = 2, window_days = 365)
```

`lookback = Inf` reproduces `mb_count_conditions(as_of = )` exactly.

## Diagnoses, and combining the two halves

`mb_extract_diagnosis()` is the ICD-10 counterpart. It needs no two-stage
filter and no prescription rule - one recorded diagnosis is the condition,
and onset is the earliest one:

```{r}
lpr <- data.frame(
  pnr      = c("1", "3"),
  C_DIAG   = c("DI10", "DI10"),
  D_INDDTO = as.Date(c("2008-01-01", "2012-01-01")),
  stringsAsFactors = FALSE
)
dx_codes <- mb_codelist(data.frame(
  condition = "hypertension", vocab_id = "ICD10", code = "I10",
  exclude = FALSE, stringsAsFactors = FALSE
))

mb_extract_diagnosis(lpr, codes = dx_codes, outdir = NULL, verbose = FALSE)
```

Combine both halves with `mb_merge_conditions()` for one condition, or
`mb_merge_all()` for every condition in a pair of output directories.
`"OR"` (the default) takes whichever date came first; `"AND"` requires both
and takes the later - Prior uses `AND` only for epilepsy. Which applies comes
from an optional `logic` column in the code list.

## Checking the code lists themselves

`C02C` (hypertension) and `C02CA` (prostate) is a real overlap in the bundled
lists - C02CA is a subgroup of C02C, so one dispensing can satisfy both:

```{r}
ov_codes <- mb_codelist(data.frame(
  condition = c("hypertension", "prostate"), vocab_id = "ATC",
  code = c("C02C", "C02CA"), exclude = FALSE, stringsAsFactors = FALSE
))
mb_overlap(ov_codes)
mb_lookup("C02CA01", ov_codes)
```

`mb_overlap()` is prefix-aware, so a condition whose code merely *contains*
another's is caught; an overlap is not automatically wrong, but one record
creating two conditions should be a decision, not a discovery. `mb_lookup()`
answers the question you want answered when a prevalence looks too high: what
would this one record be counted as.

`mb_compare()` reports what changed between two versions of a code list,
normalising both sides first so reordered rows and case differences are not
reported as changes, while rule changes (a different `window_days`) are
called out separately because they move everyone's result, not a few
people's.

## Three things that change your numbers

- **`year_min` defaults to `1997`.** That drops 1995-96 from LMDB on purpose,
  following the pipeline this package's extraction logic is derived from. Set
  `year_min = NULL` to keep every year.
- **Same-day dispensings are de-duplicated by default**, so two packages
  collected on one pharmacy visit count as one dispensing rather than
  satisfying a two-prescription rule on their own. `dedupe_same_day = FALSE`
  turns it off.
- **`window_days` requires the dispensings to fall within that many days of
  each other.** Leaving it blank (`NA`) means any distance apart. Read the
  full argument for the default of 365 days in `DECISIONS.md`.

## What is still open

Two decisions actually block trusting a number from this package - both
joint calls with Jie, because they move counts away from her existing
output, not research questions:

- **Same-day de-duplication.** Per person-day (the current default), per
  person-day-drug, or per row (no de-duplication - Jie's script's current
  behaviour)? All three are defensible and in use somewhere.
- **Prior's exclusion rules**, described above. Implement, or omit and say
  so explicitly in the paper? The rules themselves are no longer in
  question, transcribed verbatim from the Stata - only whether to apply
  them.

Also open, each blocked on one person rather than a joint call:

| Question | Blocked on |
|---|---|
| Connective tissue - confirm D86 is intended, not a typo for DD86 | Anders |
| `year_min = 1997` - why, when LMDB itself starts in 1995 | Nikolaj, or DST directly |
| How subcategories (e.g. depression/anxiety within distress) are represented | Jie |
| Optional ICD arms for KOL and allergy, alongside the ATC-only default | Jie |
| Whether a recovery option ships at all, beyond ever/never | the group |
| LPR2 psychiatric codes - Jie has working code, not yet received | Jie |
| What the package is FOR - scope, beyond which codes | the group, with Daniel |
| Where the code lives long-term, and authorship | the group |

This list is current as of the date on this vignette; `DECISIONS.md` and
`TODO.txt` in the source repository are the up-to-date versions if either has
moved since.

## Next steps

- **Verify on real register data.** Nothing in this package has touched real
  DST data yet; everything in this vignette runs on small synthetic
  examples. The next DST session checks that the code columns are the
  lengths every list assumes, that every bundled code matches something
  real, and - the architecture question - whether the window/lag rule
  behind the prescription count can run inside the database, rather than
  requiring every matching dispensing to be collected into R first.
- **Settle the two decisions above with Jie**, then the code lists can be
  unfrozen for the next round of changes.
- **Build the ICD-10 half** of the code lists, once condition naming is
  settled - three incompatible schemes are currently in play across this
  package, Jie's script, and the DST codebook, and the two halves join on
  that name, so it needs deciding before anything is written.
- **Confirm D86 with Anders**, and get Nikolaj's read on `year_min` and the
  LPR2 join.

## Where to read more

The source repository (not always shipped with the installed package) carries
the fuller documentation, each for a different audience:

- `DECISIONS.md` - every decision, its sources and what it costs
- `METHODS.md` - the same content as article-ready prose
- `ASSUMPTIONS_AND_LIMITATIONS.txt` - the short version, to read before
  reporting a number

## Reference

The bundled code lists take their starting point in the multimorbidity
algorithm published as a web supplement to:

Prior A, Fenger-Gron M, Larsen KK, Larsen FB, Robinson KM, Nielsen MG,
Christensen KS, Mercer SW, Vestergaard M. The association between perceived
stress and mortality among people with multimorbidity: a prospective
population-based cohort study. *American Journal of Epidemiology*.
2016;184(3):199-210. doi: 10.1093/aje/kwv324

regmorbidity is not an implementation of the algorithm in that supplement.
The code lists are a starting point, revised for this package's own use; see
`DECISIONS.md` and `CODELIST_REVIEW.md` in the source repository for exactly
where and why they diverge.
