---
title: "Introduction to regmorbidity"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to regmorbidity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(regmorbidity)
```

## What this vignette is for

This vignette **deepens** the README. Read the README first for install, what
you supply, the DST DuckDB path, the recommended five-step workflow, and the
core function list. This page does **not** repeat a full API inventory.

regmorbidity derives **condition indicators with dates** from Danish register
data - hospital diagnoses (LPR) and dispensings (LMDB) - using editable CSV
code lists rather than hard-coded rules. It computes no score and is not an
implementation of any published multimorbidity index. The bundled lists take
their starting point in Prior et al. (2016) and are a starting point to
revise, not an instrument to cite.

## Toy data note

The executable chunks below use **tiny synthetic** LPR / LMDB frames and
**in-memory `data.frame` code lists** for hypertension only. Those
`data.frame(...)` lists are teaching toys so the vignette runs without the
register. For real work, load the bundled CSVs with `mb_codelist()` (or pass
your own CSV folder / file). Do not copy the toy `data.frame` lists into a
study script and treat them as the package definitions.

Danish register ICD codes often carry a leading **D** (`DI10` in the register
vs `I10` in the lists). The package normalises both forms before matching, so
the toy LPR below uses `DI10` while the toy list uses `I10` on purpose.

## Worked miniature (in memory)

One hypertension example end to end: extract both halves, merge with OR,
reshape to wide, then count.

```{r}
# ---------------------------------------------------------------------------
# Tiny synthetic registers (vignette only - replace with real LPR / LMDB)
# ---------------------------------------------------------------------------
lpr <- data.frame(
  pnr      = c("1", "2", "3"),
  C_DIAG   = c("DI10", "DI10", "DE11"),   # Danish D-prefix form
  D_INDDTO = as.Date(c("2008-01-01", "2009-06-01", "2011-01-01")),
  stringsAsFactors = FALSE
)
lmdb <- data.frame(
  pnr  = c("1", "1", "2", "2", "3"),
  atc  = c("C09AA02", "C09AA02", "C09AA02", "C09AA02", "A10BA02"),
  atc2 = c("C09", "C09", "C09", "C09", "A10"),
  eksd = as.Date(c("2010-01-01", "2010-03-01",
                   "2011-01-01", "2011-04-01", "2012-01-01")),
  stringsAsFactors = FALSE
)

# ---------------------------------------------------------------------------
# Toy code lists (vignette only). Real studies: codes <- mb_codelist()
# or mb_codelist(conditions = "hypertension"), not data.frame toys.
# ---------------------------------------------------------------------------
rx_codes <- mb_codelist(data.frame(
  condition = "hypertension", vocab_id = "ATC", code = "C09",
  exclude = FALSE, min_prescriptions = 2L, window_days = 365L,
  stringsAsFactors = FALSE
))
dx_codes <- mb_codelist(data.frame(
  condition = "hypertension", vocab_id = "ICD10", code = "I10",
  exclude = FALSE, stringsAsFactors = FALSE
))

# ---------------------------------------------------------------------------
# Extract. outdir = NULL keeps results in memory (handy for toys).
# Dates below are PLACEHOLDERS - replace with your study window.
# Diagnosis: from is optional. Medication: from is REQUIRED on LMDB.
# Prefer from = as.Date("1997-01-01") on real LMDB (mother-CPR; see README).
# ---------------------------------------------------------------------------
dx <- mb_extract_diagnosis(
  lpr, codes = dx_codes, outdir = NULL,
  from = as.Date("1995-01-01"),  # inclusive start; placeholder
  to   = as.Date("2018-12-31"),  # optional; omit to leave open-ended
  verbose = FALSE
)
rx <- mb_extract_medication_batch(
  lmdb, codes = rx_codes, outdir = NULL,
  from = as.Date("1997-01-01"),  # required on LMDB
  to   = as.Date("2018-12-31"),  # optional; not required
  verbose = FALSE
)

# OR = earlier of dx / rx; AND = both required, later date.
# An optional `logic` column in the code list can set this per condition.
merged <- mb_merge_conditions(dx$hypertension, rx$hypertension, logic = "OR")

# Long form = which conditions (one row per person x condition x onset):
#   pnr | condition    | onset_date
#   1   | hypertension | 2008-01-01
#   2   | hypertension | 2009-06-01
#
# Wide = one column per condition (onset dates).
# mb_count_conditions adds n_conditions and multimorbid (>= 2). It does NOT
# re-list which conditions; those stay as the onset-date columns.
wide <- mb_to_wide(merged)
mb_count_conditions(wide, as_of = "2015-01-01")  # placeholder as-of date
```

### Example long output (shape)

After merge, the long table looks like this (illustrative rows):

| pnr | condition    | onset_date |
|-----|--------------|------------|
| 1   | hypertension | 2008-01-01 |
| 2   | hypertension | 2009-06-01 |

**Count vs which-conditions.** `mb_to_wide()` keeps *which* conditions as
onset-date columns. `mb_count_conditions(wide, as_of = )` returns that same
wide table plus `n_conditions` (how many are present as of the date) and
`multimorbid` (`TRUE` when `n_conditions >= 2`). It does not invent a separate
list of condition names.

## Directory form (real DST-style outdirs)

Same pipeline writing one `.rds` per condition. Use a **fresh** `outdir`, or
clear old files first: with `resume = TRUE` (the default), extractors skip
conditions whose `.rds` already exists. Reusing a dirty folder of leftover
results can silently keep stale onsets.

```{r, eval = FALSE}
# REPLACE placeholder dates with your study window.
# codes <- mb_codelist()   # all 36 bundled conditions
# or: codes <- mb_codelist(conditions = c("hypertension", "diabetes"))

mb_extract_diagnosis(
  lpr, codes = dx_codes, outdir = "data/dx",
  from = as.Date("1995-01-01"),
  to   = as.Date("2018-12-31")   # optional; not required
)
mb_extract_medication_batch(
  lmdb, codes = rx_codes, outdir = "data/rx",
  from = as.Date("1997-01-01"),  # required on LMDB (mother-CPR; see README)
  to   = as.Date("2018-12-31")   # optional; not required
)

# long = person x condition x onset_date (the "which conditions" table)
long <- mb_merge_all("data/dx", "data/rx")
wide <- mb_to_wide(long)
mb_count_conditions(wide, as_of = "2015-01-01")
```

Medication-only studies (no diagnosis half) can load one outdir directly:

```{r, eval = FALSE}
# Skips *_all_events.rds automatically if any are present
long_rx_only <- mb_load_conditions("data/rx")
```

## Restricting to a cohort (`ids`)

For a study cohort rather than everyone in the register, pass `ids` on the
extract call. That is a join against the lazy connection, not a written-out
id list dumped to disk.

```{r, eval = FALSE}
mb_extract_medication_batch(
  lmdb, codes = rx_codes, outdir = "data/rx",
  from = as.Date("1997-01-01"),  # required on LMDB
  to   = as.Date("2018-12-31"),  # optional; not required
  ids  = cohort$pnr              # your cohort person ids
)
```

## Batch vs sequential detail

Prefer **batch** (`mb_extract_medication_batch`) for medication onset: one SQL
pass, low RAM, same onset `.rds` shape as sequential.

Switch to **sequential** (`mb_extract_medication`) when:

1. you need **raw events** for prevalence (`keep_events = TRUE`), or
2. you want one-condition disk checkpoints / resume on an awkward cohort or
   year range.

**Why both exist.** Batch never materialises matching dispensings in R; it asks
DuckDB for the onset only - that is the right default for onset studies.
Sequential walks conditions (with an optional short `atc2` prefilter), and with
`keep_events = TRUE` it also writes `<condition>_all_events.rds` beside the
onset file. Batch has no equivalent side file, so prevalence cannot be fed from
a batch-only outdir. Merge / load / wide / count stay the same either way
because both write the same onset `.rds` shape.

## Medication rules in brief

Medication conditions usually need more than one dispensing. The code list
carries `min_prescriptions` and `window_days` (often 2 within 365 days).
Same-day dispensings are de-duplicated by default
(`dedupe_same_day = TRUE`). Leaving `window_days` as `NA` means any distance
apart.

Three settings that change counts:

- **`from` / `to`** - inclusive. On LMDB, `from` is required; prefer
  `from = as.Date("1997-01-01")` because of the mother-CPR recording change
  (Pottegard et al. 2016; see README). `to` is optional; omit it to leave the
  upper end open.
- **`dedupe_same_day`** - default `TRUE`.
- **`window_days` / `min_prescriptions`** - from the code list.

## Prevalence on raw events

`mb_count_conditions()` is ever-after from onset. Use `mb_prevalence` when you
need "still evidence within the last N days" on **raw matching records**, not
onset-only `.rds`. That requires sequential extract with `keep_events = TRUE`.

```{r}
# Sequential extract so we can keep raw events for prevalence
outdir <- tempfile()
mb_extract_medication(
  lmdb, codes = rx_codes, outdir = outdir,
  from = as.Date("1997-01-01"),  # required on LMDB
  to   = as.Date("2018-12-31"),  # optional; not required
  keep_events = TRUE, verbose = FALSE
)
disp <- readRDS(file.path(outdir, "hypertension_all_events.rds"))

mb_prevalence(
  disp, as_of = "2011-01-01", lookback = 365,
  min_prescriptions = 2, window_days = 365
)
```

- `keep_events = TRUE` writes `<condition>_all_events.rds`. Batch has no
  equivalent.
- With `outdir = NULL`, side files are not written.
- `lookback = Inf` matches `mb_count_conditions(as_of = )`.

## Merge logic (OR / AND)

`"OR"` (default) takes whichever of diagnosis / medication came first;
`"AND"` requires both and takes the later date. Which applies can come from
an optional `logic` column in the code list. `mb_merge_all()` walks a pair of
extract directories; `mb_merge_conditions()` merges one condition's two data
frames in memory (as in the miniature above).

## Optional: Prior exclusion rules (provisional)

Extraction does **not** apply Prior's cross-condition exclusions.

Use `mb_apply_exclusions` when you already have a long or wide onset table and
want an optional stage-2 pass. This is **provisional / first-pass**: incomplete
versus Prior, and not part of the recommended five-step workflow. Distress was
dropped from the default lists. Hypertension still has gaps around C03 / heart
failure / CKD.

Timing options (pass via `timing=`):

- **`"as_of"`** - default when `as_of` is given: apply exclusions relative to
  that date.
- **`"onset_order"`** - order by onset dates among conditions present.
- **`"ever_ever"`** - if both conditions ever appear, apply the exclusion
  regardless of timing.

```{r, eval = FALSE}
# long from mb_merge_all() / mb_load_conditions(), or wide from mb_to_wide()
long2 <- mb_apply_exclusions(long, as_of = as.Date("2015-01-01"))
# timing: "as_of" (default when as_of given), "onset_order", or "ever_ever"
```

## Advanced: code-list QA and inspect

These helpers are for checking lists and register columns before a long DST
run. They are not part of the recommended five-step workflow.

**`mb_check_codes`** - for each code in your list, count how many register rows
match (prefix match, same as the extractors). Typos, retired codes, or a
missing ATC chapter otherwise give **silent zeros**.

```{r}
# Worked example on the toy LMDB: each list code -> match count
mb_check_codes(lmdb, rx_codes, code_col = "atc")
```

A zero here means the extractor would also return empty for that code - fix
the list (or the register window) before a multi-hour DST run.

```{r, eval = FALSE}
mb_lookup("C02CA01")          # which conditions claim this code?
mb_overlap(vocab = "ATC")     # codes shared between conditions
```

**`mb_inspect_code_lengths`** - reports string lengths of code columns in the
*register* sample (e.g. whether `atc2` is 3 characters). Use before sequential
medication extract / prefilter debugging. This is **not** for reviewing the
code list; that is `mb_check_codes`.

Why it exists: the sequential path can push a coarse filter on `atc2` with
`prefilter_len = 3`. If `atc2` is not 3 characters (or is missing), that
prefilter silently matches nothing.

```{r}
mb_inspect_code_lengths(lmdb, cols = c("atc", "atc2"))
```

### Short failure modes

- **Wrong colnames** - defaults are LPR `pnr` / `C_DIAG` / `D_INDDTO` and LMDB
  `pnr` / `atc` / `eksd`. Pass `id_col`, `code_col`, and `date_col` when your
  tables differ.
- **Silent zeros** - codes that never match the register look like "nobody has
  this condition". Run `mb_check_codes` first.
- **Empty extract** - possible and valid when no rows qualify in the window;
  also the symptom of a bad code, bad date window, or wrong column name.
- **Dirty `outdir`** - do not reuse a folder of old `.rds` blindly; resume
  skips existing files.

## Where to read more

- **README** - install, DST path, recommended workflow, core functions, batch
  vs sequential overview.
- **`?regmorbidity`** - help for every exported function.

`ASSUMPTIONS_AND_LIMITATIONS.txt`, `DECISIONS.md`, and `METHODS.md` are
**local working-tree notes** (often gitignored). They are not part of a normal
GitHub install. If you installed from GitHub, use the README, this vignette,
and `?help`.
