---
title: "Introduction to regmorbidity"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to regmorbidity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(regmorbidity)
```

## What this package does

regmorbidity derives **condition indicators with dates** from Danish register
data - hospital diagnoses (LPR) and dispensings (LMDB) - using editable CSV
code lists rather than hard-coded rules. It computes no score and is not an
implementation of any published multimorbidity index.

The public surface is **16 exported functions**. Prefer the full-list extract
path below; use a sequential one-condition extract only when the cohort or
year range is large (see later). Downstream is one path: merge, load, wide,
count, prevalence.

Read `ASSUMPTIONS_AND_LIMITATIONS.txt` in the source repository before
reporting a number from a real run. Output is onset-only by default.

## What you supply

1. **LPR** as **one** combined table: person id, ICD code, contact date
   (defaults: `pnr`, `C_DIAG`, `D_INDDTO`). The package does not merge LPR2,
   LPR3, or psychiatric tables for you.
2. **LMDB**: person id, **full** ATC, dispensing date (defaults: `pnr`,
   `atc`, `eksd`). An optional short ATC column (e.g. `atc2`) is used only by
   the sequential medication extractor as a pushed-down prefilter.
3. **A code list** via `mb_codelist()` (bundled lists are **ATC-only**;
   diagnosis needs your ICD rows / CSV / data frame, or a filtered list).

Date windows use inclusive `from` / `to`. On LMDB medication extractors,
`from` is **required** (error if omitted/`NULL`); `to` is optional. Prefer
`from = as.Date("1997-01-01")` unless you have handled the pre-1997
mother-CPR artifact another way (see `DECISIONS.md`). Diagnosis extract
may still omit `from` (entire register).

Optional sanity checks before a long extract:

```{r, eval = FALSE}
mb_inspect_codes(lpr)           # code column lengths
mb_check_codes(lpr, codes)      # does every list code match something
mb_inspect_codes(lmdb)
mb_check_codes(lmdb, codes)
```

## Happy path

Extractors take the **full code list** and write one onset `.rds` per
condition. Diagnoses first, then medications, then merge.

```{r}
# Tiny synthetic registers for this vignette only
lpr <- data.frame(
  pnr      = c("1", "2", "3"),
  C_DIAG   = c("DI10", "DI10", "DE11"),
  D_INDDTO = as.Date(c("2008-01-01", "2009-06-01", "2011-01-01")),
  stringsAsFactors = FALSE
)
lmdb <- data.frame(
  pnr  = c("1", "1", "2", "2", "3"),
  atc  = c("C09AA02", "C09AA02", "C09AA02", "C09AA02", "A10BA02"),
  eksd = as.Date(c("2010-01-01", "2010-03-01",
                   "2011-01-01", "2011-04-01", "2012-01-01")),
  stringsAsFactors = FALSE
)

codes <- mb_codelist()
```

```{r, eval = FALSE}
# Bundled mb_codelist() is ATC-only. Diagnosis needs ICD rows (example below
# mirrors the in-memory miniature), or your own ICD CSV / filtered list.
dx_codes <- mb_codelist(data.frame(
  condition = "hypertension", vocab_id = "ICD10", code = "I10",
  exclude = FALSE, stringsAsFactors = FALSE
))
rx_codes <- mb_codelist()   # bundled ATC

mb_extract_diagnosis(
  lpr,
  codes  = dx_codes,
  outdir = "data/dx",
  from   = as.Date("1995-01-01")
)

mb_extract_medication_batch(
  lmdb,
  codes  = rx_codes,
  outdir = "data/rx",
  from   = as.Date("1997-01-01")
)

# Merge returns a long data frame. mb_load_conditions() is for an extract
# outdir when you are not merging (e.g. medication-only).
long <- mb_merge_all("data/dx", "data/rx")
wide <- mb_to_wide(long)
mb_count_conditions(wide, as_of = "2015-01-01")
```

Worked miniature (in memory, no files):

```{r}
dx_codes <- mb_codelist(data.frame(
  condition = "hypertension", vocab_id = "ICD10", code = "I10",
  exclude = FALSE, stringsAsFactors = FALSE
))
rx_codes <- mb_codelist(data.frame(
  condition = "hypertension", vocab_id = "ATC", code = "C09",
  exclude = FALSE, min_prescriptions = 2L, window_days = 365L,
  stringsAsFactors = FALSE
))

dx <- mb_extract_diagnosis(lpr, codes = dx_codes, outdir = NULL,
                           verbose = FALSE)
rx <- mb_extract_medication_batch(lmdb, codes = rx_codes, outdir = NULL,
                                  from = as.Date("1997-01-01"),
                                  verbose = FALSE)

merged <- mb_merge_conditions(dx$hypertension, rx$hypertension)
wide <- mb_to_wide(merged)
mb_count_conditions(wide, as_of = "2015-01-01")
```

For a study cohort rather than everyone in the register, pass `ids` on the
extract call. That is a join against the lazy connection, not a written-out
id list.

```{r, eval = FALSE}
mb_extract_medication_batch(lmdb, codes = codes, outdir = "data/rx",
                            from = as.Date("1997-01-01"),
                            ids = cohort$pnr)
```

## Diagnosis and merge

`mb_extract_diagnosis()` needs no prescription rule: one matching diagnosis
is the condition, and onset is the earliest date. Feed it **one** already-
combined LPR table.

`"OR"` (default) in merge takes whichever of diagnosis / medication came
first; `"AND"` requires both and takes the later date. Which applies can
come from an optional `logic` column in the code list.
`mb_merge_conditions()` merges one condition; `mb_merge_all()` walks a pair
of output directories.

## Medication rules in brief

Medication conditions usually need more than one dispensing. The code list
carries `min_prescriptions` and `window_days` (often 2 within 365 days).
Same-day dispensings are de-duplicated by default
(`dedupe_same_day = TRUE`), so two packages on one pharmacy visit count as
one event. Leaving `window_days` as `NA` means any distance apart.

Three settings that change counts:

- **`from` / `to`** - inclusive window. On LMDB medication extractors
  `from` is required; `to` optional (`NULL` = no upper bound). Prefer
  `from = as.Date("1997-01-01")` on LMDB (mother-CPR artifact).
- **`dedupe_same_day`** - default `TRUE`.
- **`window_days` / `min_prescriptions`** - from the code list, overridable
  on the extract call where the API allows.

## Ever/never, or a lookback

`mb_count_conditions()` counts a condition from its onset date on: once met,
always met (as of the date you pass). State that estimand wherever you
report a number.

If you need "still evidence within the last N days", use `mb_prevalence()`
on **raw matching records**, not onset-only `.rds`. That is one reason to
use the sequential medication extractor with `keep_events = TRUE`, which
writes `<condition>_all_events.rds` beside the onset file. Batch does not
collect raw rows. With `outdir = NULL`, side files are not written.

```{r}
outdir <- tempfile()
mb_extract_medication(lmdb, codes = rx_codes, outdir = outdir,
                      from = as.Date("1997-01-01"),
                      keep_events = TRUE, verbose = FALSE)
disp <- readRDS(file.path(outdir, "hypertension_all_events.rds"))

mb_prevalence(disp, as_of = "2011-01-01", lookback = 365,
             min_prescriptions = 2, window_days = 365)
```

`lookback = Inf` matches `mb_count_conditions(as_of = )`.
`mb_load_conditions()` skips `*_all_events.rds` automatically.

## One condition at a time

Prefer the full-list path. Switch to sequential when the population or
year range is so large that a full-list pass is awkward, or when you need
`keep_events`:

- Diagnoses: `mb_extract_diagnosis()` with a code list filtered to one
  condition (or run the full list if LPR is manageable).
- Medications: `mb_extract_medication()` (not batch). Supports
  `keep_events`; batch does not.

Both still write the same onset `.rds` shape, so merge / load / wide /
count stay unchanged.

## Code-list overlaps

Some bundled ATC ranges nest (e.g. hypertension `C02C` and prostate
`C02CA`). An overlap is not automatically wrong, but one dispensing
creating two conditions should be a deliberate choice. List them with:

```{r, eval = FALSE}
# mb_overlap() / mb_lookup() are exported (call with ::)
mb_overlap()              # all vocabs in the list
mb_overlap(vocab = "ATC") # ATC only
mb_lookup("C02CA01")                     # which conditions claim a code
```

## Optional: Prior exclusion rules (stage 2)

Extraction does **not** apply Prior's cross-condition exclusions. After you
have a long or wide onset table, you can optionally call
`mb_apply_exclusions()` (provisional; content still open with coauthors).
It is not on the happy path above.

```{r, eval = FALSE}
# long from mb_merge_all() / mb_load_conditions(), or wide from mb_to_wide()
long2 <- mb_apply_exclusions(long, as_of = as.Date("2015-01-01"))
# timing: "as_of" (default when as_of given), "onset_order", or "ever_ever"
```

With the bundled ATC-only lists this can only approximate hypertension
(C03 folded in; no heart-failure / kidney arms) and only partly cover
distress (needs emotion / schizophrenia / eating disorder for Prior
completeness). See `DECISIONS.md` 5.1.

## Where to read more

In the source repository (not always shipped with the installed package):

- `ASSUMPTIONS_AND_LIMITATIONS.txt` - short checklist before reporting a number
- `DECISIONS.md` - design choices, sources, and costs
- `METHODS.md` - article-ready prose

