---
title: "Introduction to regmorbidity"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to regmorbidity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(regmorbidity)
```

## What this package does

regmorbidity derives **condition indicators with dates** from Danish register
data — hospital diagnoses (LPR) and dispensings (LMDB) — using editable CSV
code lists rather than hard-coded rules. It computes no score and is not an
implementation of any published multimorbidity index.

The package API is a fixed set of **exported functions** (see README). How many conditions you derive depends on your code-list CSVs, not on that API count. The README is the inventory
and short one-liners for every function; this vignette owns the worked examples and the
decisions the README only points at (batch vs sequential, `keep_events`,
prevalence lookback, inspect lengths, exclusion timing).

**Use `mb_codelist` when** you need to load or filter lists before extract or
QA (`conditions=`, `vocab=`). Bundled lists are **ATC + first-pass ICD**
(distress archived out of the default set).

## What you supply

1. **LPR** as **one** combined table: person id, ICD code, contact date
   (defaults: `pnr`, `C_DIAG`, `D_INDDTO`). The package does not merge LPR2,
   LPR3, or psychiatric tables for you.
2. **LMDB**: person id, **full** ATC, dispensing date (defaults: `pnr`,
   `atc`, `eksd`). An optional short ATC column (e.g. `atc2`) is used only by
   the sequential medication extractor as a pushed-down prefilter.
3. **A code list** via `mb_codelist()` — filter with `conditions=` and/or
   `vocab=` (e.g. `vocab = "ICD10"` for diagnosis, `vocab = "ATC"` for
   medication).

Date windows use inclusive `from` / `to`. On LMDB medication extractors,
`from` is **required** (error if omitted/`NULL`); `to` is optional. Prefer
`from = as.Date("1997-01-01")` unless you have handled the pre-1997
mother-CPR artifact another way. Diagnosis extract may still omit `from`
(entire register).

## Preflight QA

**Use `mb_check_codes` when** you want an optional preflight before a long
DST extract: for each code in your list, it counts how many register rows
match (prefix match, same as the extractors). Typos, retired codes, or a
missing ATC chapter otherwise give silent zeros.

**Use `mb_lookup` when** you need to see which conditions claim a given code.

**Use `mb_overlap` when** you want codes shared between conditions (nesting
is not automatically wrong, but dual-counting should be deliberate).

```{r}
# Tiny synthetic registers for this vignette
lpr <- data.frame(
  pnr      = c("1", "2", "3"),
  C_DIAG   = c("DI10", "DI10", "DE11"),
  D_INDDTO = as.Date(c("2008-01-01", "2009-06-01", "2011-01-01")),
  stringsAsFactors = FALSE
)
lmdb <- data.frame(
  pnr  = c("1", "1", "2", "2", "3"),
  atc  = c("C09AA02", "C09AA02", "C09AA02", "C09AA02", "A10BA02"),
  atc2 = c("C09", "C09", "C09", "C09", "A10"),
  eksd = as.Date(c("2010-01-01", "2010-03-01",
                   "2011-01-01", "2011-04-01", "2012-01-01")),
  stringsAsFactors = FALSE
)

rx_codes <- mb_codelist(data.frame(
  condition = "hypertension", vocab_id = "ATC", code = "C09",
  exclude = FALSE, min_prescriptions = 2L, window_days = 365L,
  stringsAsFactors = FALSE
))
dx_codes <- mb_codelist(data.frame(
  condition = "hypertension", vocab_id = "ICD10", code = "I10",
  exclude = FALSE, stringsAsFactors = FALSE
))

# Worked example: each list code → match count in the register
mb_check_codes(lmdb, rx_codes, code_col = "atc")
```

A zero here means the extractor would also return empty for that code — fix
the list (or the register window) before a multi-hour DST run.

```{r, eval = FALSE}
mb_lookup("C02CA01")          # which conditions claim this code?
mb_overlap(vocab = "ATC")     # codes shared between conditions
```

### Advanced: column string lengths in the register

**Use `mb_inspect_code_lengths` when** you are about to run the *sequential*
medication extractor with a short prefilter column and need to confirm that
column’s string lengths in the *register* (sample) — e.g. whether `atc2` is
really 3 characters. This is **not** for reviewing the code list; that is
`mb_check_codes`.

Why it exists: the sequential path can push a coarse filter on `atc2` with
`prefilter_len = 3`. If `atc2` is not 3 characters (or is missing), that
prefilter silently matches nothing.

```{r}
mb_inspect_code_lengths(lmdb, cols = c("atc", "atc2"))
```

## Worked miniature (in memory)

**Use `mb_extract_diagnosis` when** you have LPR and want onset from one
matching diagnosis.

**Use `mb_extract_medication_batch` when** you want medication onset for many
conditions in one low-RAM SQL pass (preferred path; no raw events kept).

**Use `mb_merge_conditions` when** you already have two in-memory halves for
*one* condition (vignette / debug). **Use `mb_merge_all` when** you have two
extract directories from a real run.

**Use `mb_to_wide` when** you need one row per person and one onset-date
column per condition. **Use `mb_count_conditions` when** you want ever-after
counts as of a date (once met, always met).

```{r}
dx <- mb_extract_diagnosis(lpr, codes = dx_codes, outdir = NULL,
                           verbose = FALSE)
rx <- mb_extract_medication_batch(lmdb, codes = rx_codes, outdir = NULL,
                                  from = as.Date("1997-01-01"),
                                  verbose = FALSE)

# OR = earlier of dx/rx; AND = both required, later date.
# An optional `logic` column in the code list can set this per condition.
merged <- mb_merge_conditions(dx$hypertension, rx$hypertension, logic = "OR")
wide <- mb_to_wide(merged)
mb_count_conditions(wide, as_of = "2015-01-01")
```

Directory form of the same pipeline (not evaluated here):

```{r, eval = FALSE}
mb_extract_diagnosis(lpr, codes = dx_codes, outdir = "data/dx",
                     from = as.Date("1995-01-01"))
mb_extract_medication_batch(lmdb, codes = rx_codes, outdir = "data/rx",
                            from = as.Date("1997-01-01"))
long <- mb_merge_all("data/dx", "data/rx")
wide <- mb_to_wide(long)
mb_count_conditions(wide, as_of = "2015-01-01")
```

For a study cohort rather than everyone in the register, pass `ids` on the
extract call (a join against the lazy connection, not a written-out id list):

```{r, eval = FALSE}
mb_extract_medication_batch(lmdb, codes = rx_codes, outdir = "data/rx",
                            from = as.Date("1997-01-01"),
                            ids = cohort$pnr)
```

**Use `mb_load_conditions` when** you have one extract outdir and are not
merging (e.g. medication-only). It skips `*_all_events.rds` automatically.

```{r, eval = FALSE}
long_rx_only <- mb_load_conditions("data/rx")
```

## Batch vs sequential (and `keep_events`)

Prefer **batch** for medication onset: one SQL pass, low RAM, same onset
`.rds` shape as sequential.

Switch to **sequential** (`mb_extract_medication`) when:

1. you need **raw events** for prevalence (`keep_events = TRUE`), or
2. you want one-condition disk checkpoints / resume on a awkward cohort or
   year range.

**Why the RAM story matters:** batch never materialises matching dispensings
in R; it asks DuckDB for the onset only. Sequential walks conditions (with
optional prefilter on `atc2`), and with `keep_events = TRUE` it also writes
`<condition>_all_events.rds` beside the onset file. Batch has no equivalent
side file — so prevalence cannot be fed from a batch-only outdir.

Both write the same onset `.rds` shape, so merge / load / wide / count stay
unchanged.

## Medication rules in brief

Medication conditions usually need more than one dispensing. The code list
carries `min_prescriptions` and `window_days` (often 2 within 365 days).
Same-day dispensings are de-duplicated by default
(`dedupe_same_day = TRUE`). Leaving `window_days` as `NA` means any distance
apart.

Three settings that change counts:

- **`from` / `to`** — inclusive. On LMDB, `from` is required; prefer
  `from = as.Date("1997-01-01")`.
- **`dedupe_same_day`** — default `TRUE`.
- **`window_days` / `min_prescriptions`** — from the code list.

## Prevalence on raw events

`mb_count_conditions()` is ever-after from onset. **Use `mb_prevalence` when**
you need "still evidence within the last N days" on **raw matching records**,
not onset-only `.rds`.

```{r}
outdir <- tempfile()
mb_extract_medication(lmdb, codes = rx_codes, outdir = outdir,
                      from = as.Date("1997-01-01"),
                      keep_events = TRUE, verbose = FALSE)
disp <- readRDS(file.path(outdir, "hypertension_all_events.rds"))

mb_prevalence(disp, as_of = "2011-01-01", lookback = 365,
             min_prescriptions = 2, window_days = 365)
```

- `keep_events = TRUE` writes `<condition>_all_events.rds`. Batch has no
  equivalent.
- With `outdir = NULL`, side files are not written.
- `lookback = Inf` matches `mb_count_conditions(as_of = )`.

## Merge logic (OR / AND)

`"OR"` (default) takes whichever of diagnosis / medication came first;
`"AND"` requires both and takes the later date. Which applies can come from
an optional `logic` column in the code list. `mb_merge_all()` walks a pair of
extract directories; `mb_merge_conditions()` merges one condition’s two data
frames in memory (as in the miniature above).

## Optional: Prior exclusion rules (provisional)

Extraction does **not** apply Prior's cross-condition exclusions.

**Use `mb_apply_exclusions` when** you already have a long or wide onset
table and want an optional stage-2 pass — provisional; incomplete vs Prior;
not part of the recommended five-step workflow. Distress was dropped from the default lists.
Hypertension still has gaps around C03 / heart failure / CKD.

Timing options (pass via `timing=`):

- **`"as_of"`** — default when `as_of` is given: apply exclusions relative to
  that date.
- **`"onset_order"`** — order by onset dates among conditions present.
- **`"ever_ever"`** — if both conditions ever appear, apply the exclusion
  regardless of timing.

```{r, eval = FALSE}
# long from mb_merge_all() / mb_load_conditions(), or wide from mb_to_wide()
long2 <- mb_apply_exclusions(long, as_of = as.Date("2015-01-01"))
# timing: "as_of" (default when as_of given), "onset_order", or "ever_ever"
```

## Where to read more

`?regmorbidity` and the README — every exported function as a one-liner.

In the source working tree if present (often gitignored / local):

- `ASSUMPTIONS_AND_LIMITATIONS.txt` — short checklist before reporting a number
- `DECISIONS.md` — design choices, sources, and costs
- `METHODS.md` — article-ready prose
